---
title: "Everything about sample size analysis in clinical studies"
output: html_notebook
Author: Xulong Wang (emailofx@gmail.com)
---

## Test statistics of the hypothesis tests 

1. continuous outcomes: differences of the measurement average 
2. time to event outcomes: differences of expected time to events 
3. binary outcomes: differences of case proportions 

## A basic formula for sample sizes

$n = 2\sigma^2 \frac{\xi_{1-\alpha/2} + \xi_{1-\beta}}{\Delta_A}$

$\sigma^2$: variance of the outcome measures
$\Delta_A$: mean difference of the outcome measures under the alternative hypothesis
$\xi_{1-\alpha/2}$: $1-\alpha/2$ quantile of the standard normal distribution
$\xi_{1-\beta}$: $1-\beta$ quantile of the standard normal distribution

$\alpha$: the type I error, is the probability of rejecting the null hypotheisis when the null hypothesis is true
$\beta$: type II error, is the probability of fail to reject the null hypothesis when the alternative hypothesis is true

## Assumption of the basic formula

1. normality of the test statistics
2. known and equal variances in the two treatment groups
3. equal sample sizes in the two treatment groups
4. iid observations

## Survival data

Point: For survival outcomes, event number determines statistical power, because event number determines the test statistics
Thus, it also determines the sample size analysis results

> Why:
Test statistics of survival analysis, difference of expected time to failure, are transformed to ratio of hazard rate (HR) in each group
Turns out, with survival data generated from the same model (thus the same hazard rate), hazard rate estimates are the same for two sample data under two situations: (1) follow k samples until all have failed; (2) follow a larger number of samples (n) until k have failed
> More details in section: Nitty-gritty of the exponential distribution

## Sample size analysis procedure 

1. calculate event number: taking $\alpha$, $\beta$, $HR_A$, and patient proportion 
2. calculate final sample size: taking baseline event rate, drop-out rate, recruitment schedule and follow-up schedule

> If the total study period (ST) is divided into the accrual (AC) and follow-up interval (FU): ST = AC + FU
> Here are general rules on how accrual and follow-up intervals affect the required patient number
> The rule of thumb is: longer average on-study time leads to smaller patient number
> If the total study period is fixed, longer follow-up (shorter accrual) means longer average on-study time
> But this requires higher accrual rate, which might be a bottle-neck and might be fixed
> With accrual rate fixed: 
> Increase the accural time leads to more total patient number, shorter follow-up time, and shorter total study time 
> Decrease the accural time leads to less total patient number, longer follow-up time, and longer total study time

## Test using the GSDESIGN package

```{r}
library(gsDesign)

y1 = nSurv(hr0 = 1, hr = 0.5, lambdaC = 1/5, alpha = 0.05, beta = 0.2, eta = 0.1, gamma = 20, R = 7, T = 20)
y1$n
y1$T

y2 = nSurv(hr0 = 1, hr = 0.5, lambdaC = 1/5, alpha = 0.05, beta = 0.2, eta = 0.1, gamma = 20, R = 5, T = 50)
y2$n
y2$T

```

## Nitty-gritty of the exponential model 

Exponential distribution model the distribution of the time until next event occurs
Formal definition of exponential distribution: the probability distribution of the time *between* the events in a Poisson process

$T \sim Exp(\lambda)$
$\lambda$: event rate, or number of events per unit time

> simulations address:
> the same inference if follow d subjects until all have failed or if follow some larger number n until d have failed

```{r}
n = 3
lambda = 3

t1 = rexp(1e3, rate = n * lambda)
hist(t1, breaks = 1e2)

# Let x ~ Exp(lambda)
# x_1 ~ Exp(n * lambda)

x1 = rexp(n = 1e3, rate = 1)
x2 = rexp(n = 1e3, rate = 10)

hist(x1)
hist(x2)

mean(x1)
mean(x2)

t1 = sapply(1:1e3, function(x) min(rexp(n = 10, rate = 1)))
hist(t1)
mean(t1)

# delta method for variance-stabilizing

x1 = rexp(n = 1e4, rate = 2)
mean(x1)
var(x1)

xh = sapply(1:1e3, function(x) mean(rexp(n = 10, rate = 2)))
mean(xh)
var(xh)

xh2 = sapply(1:1e4, function(x) log(mean(rexp(n = 10, rate = 2))))
mean(xh2)
-log(2)
var(xh2)
```

## Questions

> What is competing risk?  
