# Survival analysis
> Survival analysis refers to the modeling of time to event data

# Nomenclatures

The probability that a subject survives longer than time t
Pr(T > t) = S(t)
S(t) is the survival function

The probability that a subject survives up until time t
Pr(T <= t) = F(t) = 1 - S(t)
F(t) is the cumulative distribution of the event density function f(t)
f(t) reflects the probability of event at a given time
f(t) = dF(t) / dt = lim { Pr(t <= T < t + dt) / dt }

The probability of event at a given time, conditional on the subject still survives at the given time
h(t) = f(t) / S(t) or  f(t) = h(t) * S(t)
h(t) is the harzard function

# Deductions

h(t) = F’(t) / S(t) = -S’(t) / S(t) = -d {ln(S(t))} / dt

# Likelihood function to fit parameters (p)

For uncensored data: Pr(T = Ti | p) = f(Ti | p)
For right-censored data: Pr(T > Ti | p) = S(Ti | p) = 1 - F(Ti | p)

Likelihood function L(p) is the product of probability function of each data point

# Exponential model

The exponential model of survival data assume exponential distribution for the survival function

$$S(t) = exp(-\lambda t)$$
$$h(t) = \lambda$$
$$f(t) = \lambda * exp(-\lambda t)$$
# Event density function

```{r}
k = 2
t = 1:1e3 / 1e2 
y = k * exp(-k * t)
plot(t, y)
```

# Likelihood function

```{r}
h = 1:1e3 / 1e2
l1 = h^10 * exp(-h * 10)
l2 = h^10 * exp(-h * 15)

which.max(l1)
which.max(l2)

plot.new()
frame()
plot(h, l1); lines(h, l2, col = "red")
```

# Weibull model

Exponential model has only 1 parameter (scale parameter).
Generalization of the exponential model to include a shape parameter is the Weibull model.

$$S(t) = exp(-\lambda t^\gamma)$$
$$h(t) = \lambda * \gamma * t^{\gamma - 1}$$
$$f(t) = \lambda * \gamma * t^{\gamma - 1}exp(-\lambda t^\gamma)$$

# Proportional hazards model

It is the usual practice to model $\lambda$ by other exploratory factors using linear regression models (covariates)

$$log(\lambda) = \alpha + \beta * x$$

$$h(t) = exp(\alpha + \beta * x) * \gamma * t^{\gamma - 1}$$
By modeling $log(\lambda)$ using linear regression models, the ratio of the hazard functions of two different values of x ($x_1$ and $x_2$) equals $exp{\beta (x_1 - x_2)}$.

This ratio is a constant proportion that depends only on the covariate and not on time. Hazard ratio between two groups remains constant over time.

>> Thus it is called a proportional hazards model.

# Cox PH model

Cox introduced a model of the hazard function that follows: $$h(t | x) = h_0(t) * exp(\beta x)$$

In this model, $h_0(t)$ is called baseline hazard function, in terms of the hazard function when x equals 0. There are two key aspects of the baseline hazard function $h_0(t)$: 

1. $h_0(t)$ is still a time-dependent function. However, Cox model does not assume its specific form. 
2. $h_0(t)$ is independent of x. x was modeled into $exp(\beta x)$

In a nutshell, hazard rate is modeled as multiple of baseline hazard rate, which is a time-dependent function assuming all covariates are zero.

Cox model can be viewed as a special case of the Weibull model. And obviously, Cox model is also a proportional hazard model.

In a nutshell, Cox PH model is a linear model for the log of the hazard ratio. As a comparison, Weibull model is a direct model of the hazard rate, for a given covariate value at a given time.

Cox proposed partial maximal likelihood method to estimate the model parameters. 

## How the partial maximal likelihood method was done? 

# Kaplan Meier curve and Logrank test

# How are the p-values look like for Logrank test, Weibull modek, and Cox model
