> http://www.drizopoulos.com/courses/emc/ep03_%20survival%20analysis%20in%20r%20companion

To be continued ...

## Cox PH model

Baseline hazard refers to the hazard rate when the covariate is set to its baseline, i.e. 0 for numerical covariate. 
This may not be confused with the baseline time.
Proportional Hazard (PH) model assumes that: over the time, hazard ratio is constant for two different values of the covariate.

## Stratified PH model

The stratified model *allows* different baseline hazard function for each stratum of the stratification factor.
Meanwhile, effect of the stratification factor, in term of hazard ratio, is not estimated.

This is a remedy to deal with covariates of non-proportional effects, which violates the PH model assumption and therefore not appropriate as model covariates.

Note that while stratified Cox PH model *allows* different baseline hazard functions of different strata, Cox PH model in general does not estimate the baseline functions. 
Cox PH only estimates hazard ratio of the model covariates.

Baseline hazard function can be estimated by using survfit(xx) function, while xx can be a stratified Cox PH model output

```{r, include = F}
xx = coxph(OS ~ A + strata(B))
survfit(xx)
plot(survfit(xx))
```

## Clinical Trials Survival Analysis - Execution 

```{r}
library("survival")
data(lung, stanford2, package = "survival")

lung <- lung %>% mutate(sex = factor(sex, levels = 1:2, labels = c("male", "female")))
lung <- lung %>% filter(complete.cases(time, status, sex, age, ph.karno))
```

Function::Surv()
> This is to specify the type of survival data, namely, right censored, left censored, interval censored.

Function::survfit()
> This is to calculate the Kaplan-Meler estimate.

```{r}
KMfit = survfit(Surv(time, status) ~ 1, data = stanford2)

KMfit # basic outputs include: number of subjects, number of events, median survival time and its 95% confidence interval

summary(KMfit, times = c(0, 1e3, 2e3)) # use summary method and its time arguments, we obtain survival probability at specific times

quantile(KMfit, probs = 1 - c(1, 0.5, 0.25)) 
# use quantile method and its probs arguments, we obtain corresponding follow-up times at which the survival probability takes specific values
# we also obtain the lower and upper limits of the 95% confidence intervals for the quantiles we asked
# we specify one minus target survival probability because the quantile method works under the cumulative distribution function convention
# and CDF equals one minus survival probability

plot(KMfit, xlab = "Time to Death (days)", ylab = "Survival Probability", main = "Kaplan-Meier Estimate of S(t) for the Stanford Data")
```

## Statistical Tests

Hypothesis testing of *one* grouping variable, i.e. test whether survival functions of different groups of subjects differ significantly

```{r}
logrank = survdiff(Surv(time, status == 2) ~ sex, data = lung) # log-rank test: the first and the most widely used test
petopeto = survdiff(Surv(time, status == 2) ~ sex, data = lung, rho = 1) # Peto & Peto Gehan-Wilcoxon test

# log-rank test is the most powerful test when proportional hazard (PH) assumption is satisfied
# we can plot the cumulative KM hazard function for the two groups to test the PH assumption
# when PH is satisfied, the two curves will be proportional to each other, i.e. steadily grow away from each other

KMfit = survfit(Surv(time, status == 2) ~ sex, data = lung)

plot(KMfit)
plot(KMfit, fun = "cumhaz") # cumulative hazards plot
```

## How to understand cumulative hazards?

## Accelerated Failure Time Models

Function::survfit() in the survival package fits AFT models

```{r}
expfit = survreg(Surv(time, status == 2) ~ sex, data = lung, dist = "exponential") # dist argument specifies distribution of the survival time 
weibullfit = survreg(Surv(time, status == 2) ~ sex, data = lung) # default is the Weibull distribution

lognormfit = survreg(Surv(time, status == 2) ~ sex, data = lung, dist = "lognormal") # log-normal distribution
loglogisfit = survreg(Surv(time, status == 2) ~ sex, data = lung, dist = "loglogistic") # log-logistic distribution
```

## Effects plot to visualize complex interactive models

```{r}
expand.grid() # create a dataset of all possible combinations of the model covariate
predict() # generate model predictions (in the log survival time scale) and the corresponding standard errors
lattice::xyplot() # make the plot
```

## Hypothesis tests

> when we want to test whether we can drop all the complex terms
> A statistical comparison between two nested models can be performed with a likelihood ratio test by funtion anova()

```{r}
fit_alt <- survreg(Surv(years, status2) ~ (drug + sex) * (age + I(age^2)), data = pbc2.id)
fit_null <- survreg(Surv(years, status2) ~ drug + sex + age, data = pbc2.id)
anova(fit_null, fit_alt)
```

## Residules

> we can use residules of the model to check whether the assumed distribution of survival times fits the observed data sufficiently well

## Cox PH model

