---
title: "Models to Handle Treatment Switches in Clinical Trials"
output: html_notebook
---

```{r, include = F}
library(survival)
library(survminer)
library(rpsftm)
library(ipcwswitch)
```

# Rank-preserving structural failed time model (RPSFTM package)

Relate the individual patient's observed event time T to the event time U that would have been observed had no active treatment been given.

$T_i = T^{off}_i + T^{on}_i$
$U_i = T^{off}_i + T^{on}_iexp(\psi)$
$\psi$ represents the hazard ratio

The best estimate of $\psi$ should have the same observed time to event in the absence of any treatment.
The time with no treatment is known as couter-factual time.

```{r}
with(immdef, table(def, imm))
survfit(Surv(progyrs, prog) ~ imm, data = immdef)
coxph(Surv(progyrs, prog) ~ imm, data = immdef)
noadj = survfit(Surv(progyrs, prog) ~ imm, data = immdef)
plot(noadj, main = "KM w/o adjustment", xlab = "Time", ylab = "Survival", lwd = 2, col = 1:3)

immdef$rx = 1 - immdef$xoyrs / immdef$progyrs
rpsftm_fit1 = rpsftm(formula = Surv(progyrs, prog) ~ rand(imm, rx), data = immdef) # w/o reconsoring
rpsftm_fit2 = rpsftm(formula = Surv(progyrs, prog) ~ rand(imm, rx), data = immdef, censor_time = censyrs) # w recensoring
str(rpsftm_fit1)
rpsftm_fit1$fit

rpsftm_fit1$psi
rpsftm_fit2$psi

immdef$c0 = Surv(immdef$progyrs, immdef$prog)
immdef$c1 = rpsftm_fit1$Sstar
immdef$c2 = rpsftm_fit2$Sstar

sum(immdef$prog[immdef$imm == 1]) # more censor, less event
sum(immdef$prog[immdef$imm == 0]) # less censor, more event

x1 = immdef %>% filter(imm == "0")
x2 = immdef %>% filter(imm == "1")

ggsurvplot(survfit(c0 ~ imm, data = immdef), risk.table = T, risk.table.height = 0.3, palette = c("red", "blue"), xlab = "Time")
ggsurvplot(survfit(c1 ~ imm, data = immdef), risk.table = T, risk.table.height = 0.3, palette = c("red", "blue"), xlab = "Time")
ggsurvplot(survfit(c2 ~ imm, data = immdef), risk.table = T, risk.table.height = 0.3, palette = c("red", "blue"), xlab = "Time")

table(grepl("\\+", immdef$c0), immdef$imm) # more censored in the treatment groups
table(grepl("\\+", immdef$c1), immdef$imm) # same thing, which is a problem
table(grepl("\\+", immdef$c2), immdef$imm) # now balanced
```

## How to understand the recensoring process in the RPSFTM algorithm?

>> this is a post-hoc fix to address the correlation between censor time and treatment groups
>> it works well, but I doubt whether there is any perfect mathematical proof
>> revisit this

## Inverse-probability of censoring weighting and IPCWSWITCH package

>> IPCW represents an approach for adjusting estiamtes of a treatment effect in the presence of any types of informative censoring. 
>> In the context of treatment switching, patients are artifically censored at the time of switch, and remaining observations are weighted based up on covariate values and a model of the probability of being censored. 
>> This allows patients who have not been artifically censored to be weighted in order to reflect their similarities to patients who have been censored in an attempt to remove the selection bias caused by the censoring.

>> At any given time t, the weights are defined as the inverse of the probability of being uncensored, e.g. patients who are similar with those who switch get more weights, as a compensation of those who switched and censored
>> Case weights are treated as replication weights, i.e., a case weight of 2 is equivalent to having 2 copies of that subject's observation. 

## Make the toy data

```{r}
toy = toydata; toy

toy2 = lapply(unique(toy$id), function(pid) { cat(pid, "\n")
  y1 = toy %>% filter(id == pid) %>% as.matrix
  d1 = y1[1, c("randt", "dt2", "dt3")] 
  td = y1[1, c("ps1", "ps2", "ps3")]; names(td) = c("randt", "dt2", "dt3")
  
  td = td[! is.na(d1)]; d1 = d1[! is.na(d1)]
  d1 = d1[! duplicated(td)]; td = td[! duplicated(td)]
  d1 = c(d1, y1[1, "lastdt"])
  
  y2 = data.frame(id = pid, tstart = d1[1:(length(d1)-1)], tstop = d1[-1])
  y2$TD = td[rownames(y2)]
  y2$event = 0
  y2$event[nrow(y2)] = y1[1, "status"]
  y2
}); toy2 = do.call(rbind, toy2)

toy2 = merge(toy2, toy[c("id", "age", "arm", "swtrtdt")], by = "id", all.x = T)
toy2

toy2$cens = as.numeric(! is.na(toy2$swtrtdt))
toy2$tstart = as.Date(toy2$tstart)
toy2$tstop = as.Date(toy2$tstop)

toy3 = lapply(unique(toy2$id), function(pid) {
  y1 = toy2 %>% filter(id == pid)
  ref = min(y1$tstart)
  y1$tstart = y1$tstart = ref; y1$tstop = y1$tstop - ref; y1$swtrtdt = y1$swtrtdt - ref
  y1
}); toy3 = do.call(rbind, toy3)

toy3
```

## final data and weights

```{r}
toy4 = openxlsx::read.xlsx("ipcwswitch.xlsx")
toy4 = rbind(toy4, toy4)

toy5 = ipcw(toy4, id = "id", tstart = tstart, tstop = tstop, cens = cens, arm = "arm", bas.cov = "age", conf = "TD", trunc = 0.01, type = "kaplan-meier")

```
